---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# xdtkit

<!-- badges: start -->
<!-- badges: end -->

The goal of xdtkit is to make available functions for predicting AADT (annual average daily traffic) on the Norwegian road network.

## Installation

You can install the development version of xdtkit from GitHub with:

``` r
# install.packages("pak")
pak::pak("trafikkdata/xdtkit")
```

## Example

The AADT modelling process requires some fairly large datasets that are not publicly available and are too large to include with the package. To show how the package should be used, we have included data from only Buskerud.

Code for running the full AADT model can be found in the repository [xdt-modelling](https://github.com/trafikkdata/xdt-modelling).


### Data preparation

We start by loading the package and specifying the year for the data. 

```{r}
library(xdtkit)
year <- 2025
```

Next, we will load and prepare all the data, and compute the adjacency matrix used in the INLA-step and the traffic link clusters used in the balancing step.

We first load the Buskerud traffic links, these are included in the R package.
```{r}
# Traffic links: Load and preprocess
preprocessed_traffic_links <- preprocess_traffic_links(buskerud_directed_traffic_links, year = year)
```

We also join the bus counts to the data set that contains information about which bus stops are connected to which traffic links, so that for the traffic links that have only public transport lanes, we can use the average number of buses that stops there as the AADT value. There are `r nrow(stops_on_traffic_links)` such traffic links in all of Norway in `r year`, though in Buskerud there is only `r sum(buskerud_directed_traffic_links$hasOnlyPublicTransportLanes, na.rm = TRUE)`. 

```{r}
# Bus data: Load and preprocess
bus_aadt <- calculate_bus_aadt(stops_on_traffic_links, bus_counts, year = year)
```

Notice that many of the columns in the directed traffic link data set are missing values. 
```{r}
missing_counts <- colSums(is.na(preprocessed_traffic_links))
missing_counts[missing_counts > 0]
```

For the INLA model, any missing covariate values will be replaced by 0, which is often not a good idea. So instead we fill in the missing values in a few different ways, depending on the type of variable. For the categorical variables, we create an "unknown" category, as these are often missing in similar places (private or municipal roads). For the logical column `hasPublicTransportLanes`, we assume that missing entries do not have only public transport lanes. For the numerical columns, we fill missing values by the median for that column. In addition, we remove any negative values and treat them as if they don't have data, and add `logLastYear` as a transformed variable, and we join in the bus data.

```{r}
# Fill missing values and add bus data
prepared_traffic_links <- fill_missing_values(
  df = preprocessed_traffic_links,
  unknown_impute_columns = c("functionClass", "highestSpeedLimit", "lowestSpeedLimit","maxLanes", "minLanes"),
  mode_impute_columns = c("hasOnlyPublicTransportLanes"),
  median_impute_columns = c("lastYearAadt_aadt", "lastYearAadt_heavyRatio", 
                            "lastYearAadt_heavyAadt")) |>
  remove_negative_aadt() |> 
  add_logLastYear() |>
  join_bus_to_traffic(bus_aadt)

missing_counts <- colSums(is.na(prepared_traffic_links))
missing_counts[missing_counts > 0]
```

Now only non-covariate columns are missing values.


Now we can build the adjacensy matrix. This is built based on the traffic link data alone, since that data set has columns indicating the start- and end-nodes. Using that information, we can identify the neighboring traffic links.
```{r}
# Adjacency matrix (may take several minutes to run)
adjacency_matrix <- build_adjacency_matrix(
  prepared_traffic_links,
  exclude_public_transport = TRUE)
```

We also calculate the balancing clusters. These are sub-groups of the traffic links that are separated in the road network by traffic links that have registration points. In this case, the number of traffic links in Buskerud is so small that we could have just dropped this step, but when running the model on all of Norway we need to separate into these clusters to be able to run it at all (because of memory limitations). So in this case, there will probably not be many sub-clusters, but we include it in order to illustrate what the full workflow will look like.
```{r}
# Balancing clusters
clusters <- strategic_network_clustering(
  data = prepared_traffic_links,
  year = year, 
  boundary_links = c("Trafikkdata_continuous"))
```

Finally, we load the traffic nodes for Buskerud, and add information on which nodes are not balanceable, that is, which nodes do we not want to enforce the condition that the number of incoming cars is equal to the number of outgoing cars. Importantly, the node data contains information about the legal turning movements from every incoming traffic link to that node. All of this is used in the balancing step.
```{r}
# Nodes: Load and preprocess (may take a minute to run)
nodes <- identify_unbalanceable_nodes(buskerud_nodes, prepared_traffic_links)
```


### Inla model

Now that we have our data and adjacency matrix, we can run the INLA model. 

```{r}
covariates <- ~ functionalRoadClass:maxLanes +
  functionalRoadClass:roadCategory +
  minLanes:roadCategory + functionalRoadClass +
  maxLanes + roadCategory +
  hasOnlyPublicTransportLanes + 
  functionalRoadClass*isRamp +
  lastYearAadt_logAadt

inla_model_total <- fit_inla_model(
  data = prepared_traffic_links,
  adjacency_matrix,
  fixed_effects = covariates,
  iid_effects = "roadSystem",
  family = "poisson")

inla_model_total
```

We join the predictions and their associated uncertainty to the traffic link data.
```{r}
predictions_total <- dplyr::full_join(prepared_traffic_links, inla_model_total$predictions)
```


### Balancing model

Next, we balance the INLA predictions to ensure correct traffic flow, and add the balanced results to the traffic links.

```{r}
balanced_model_total <- balance_predictions(data = predictions_total,
                                            nodes = nodes,
                                            balancing_grouping_variable = clusters,
                                            nodes_to_balance = "complete nodes",
                                            year = year)

predictions_total <- dplyr::full_join(predictions_total, balanced_model_total$balanced_res)
```

And that completes the modelling workflow!

### Examining the results

```{r}
plot_predictions_against_last_year(predictions_total)
```

```{r}
plot_prediction_relative_uncertainty_histogram(predictions_total, cap_cv = 1)
```


