---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# xdtkit

<!-- badges: start -->
<!-- badges: end -->

The goal of xdtkit is to make available functions for predicting AADT (annual average daily traffic) on the Norwegian road network.

## Installation

You can install the development version of xdtkit from GitHub with:

``` r
# install.packages("pak")
pak::pak("trafikkdata/xdtkit")
```

## Example

The AADT modelling process requires some fairly large datasets that are not publicly available and are too large to include with the package. To show how the package should be used, we have included data from (some county).

**To come: Example of AADT-process on subset of Norway.**

Until then, code for running the full AADT model can be found in the repository [xdt-modelling](https://github.com/trafikkdata/xdt-modelling).


```{r}
library(xdtkit)

year <- 2025

# Traffic links: Load and preprocess
preprocessed_traffic_links <- preprocess_traffic_links(buskerud_directed_traffic_links, year = year)

missing_counts <- colSums(is.na(preprocessed_traffic_links))
missing_counts[missing_counts > 0]


# Bus data: Load and preprocess
bus_aadt <- calculate_bus_aadt(stops_on_traffic_links, bus_counts, year = year)


# Fill missing values and add bus data
prepared_traffic_links <- fill_missing_values(
  df = preprocessed_traffic_links,
  unknown_impute_columns = c("functionClass", "highestSpeedLimit", "lowestSpeedLimit","maxLanes", "minLanes"),
  mode_impute_columns = c("hasOnlyPublicTransportLanes"),
  median_impute_columns = c("lastYearAadt_aadt", "lastYearAadt_heavyRatio", 
                            "lastYearAadt_heavyAadt")) |>
  remove_negative_aadt() |> 
  add_logLastYear() |>
  join_bus_to_traffic(bus_aadt)

missing_counts <- colSums(is.na(prepared_traffic_links))
missing_counts[missing_counts > 0]


# Nodes: Load and preprocess (may take a minute to run)
nodes <- identify_unbalanceable_nodes(buskerud_nodes, prepared_traffic_links) 


# Adjacency matrix (may take several minutes to run)
adjacency_matrix <- build_adjacency_matrix(
  prepared_traffic_links,
  exclude_public_transport = TRUE) 


# Balancing clusters
clusters <- strategic_network_clustering(
  data = prepared_traffic_links,
  year = year, 
  boundary_links = c("Trafikkdata_continuous"))


# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Model setup
# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

covariates <- ~ functionalRoadClass:maxLanes +
  functionalRoadClass:roadCategory +
  minLanes:roadCategory + functionalRoadClass +
  maxLanes + roadCategory +
  hasOnlyPublicTransportLanes + 
  functionalRoadClass*isRamp

# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# 2.a Run INLA model for total AADT.
# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

covariates_total <- update(covariates, ~ . + lastYearAadt_logAadt)

inla_model_total <- fit_inla_model(
  data = prepared_traffic_links,
  adjacency_matrix,
  fixed_effects = covariates_total,
  iid_effects = "roadSystem",
  family = "poisson")

inla_model_total

predictions_total <- dplyr::full_join(prepared_traffic_links, inla_model_total$predictions)

# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# 2.b Run balancing for total AADT.
# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
balanced_model_total <- balance_predictions(data = predictions_total,
                                            nodes = nodes,
                                            balancing_grouping_variable = clusters,
                                            nodes_to_balance = "complete nodes",
                                            year = year)

predictions_total <- dplyr::full_join(predictions_total, balanced_model_total$balanced_res)
```

